---
title: "NHATS_NCOS_datasets"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2025-03-17"
---

1st: NHATS datasets (caregivers/digital tools)

# Focusing on Digital tools first:

```{r}
# Load necessary libraries
library(haven)    # For reading SAS files
library(dplyr)    # For data manipulation
library(readr)    # For exporting CSV
library(tidyverse)

# Set file paths (Update these based on your file locations)
tracker_data <- read_sas("~/desktop/Digital-tools_caregivers/NHATS_R13_Final_Release_SAS/Tracker_files/NHATS_Round_13_Tracker_File.sas7bdat")
tab_act <- read_sas("~/desktop/Digital-tools_caregivers/NHATS_R13_Final_Release_SAS/Tab_act_files/NHATS_Round_13_Tab_Act_File.sas7bdat")

# Tab_Act => Covers assistive devices, telehealth, vision, hearing, mobility

# Save as CSV
# write.csv(tracker_data, "tracker_data.csv", row.names = FALSE)
# write.csv(tab_act, "tab_act.csv", row.names = FALSE)
```

```{r}
#Exploratory data analysis:

# Display all column names
colnames(tracker_data)
colnames(tab_act)

#Check Data Structure & Summary
# View first few rows
head(tracker_data)
head(tab_act)

# Summary statistics for numerical variables
summary(tracker_data)
summary(tab_act)


# Check structure of the datasets
str(tracker_data)
str(tab_act)

# Count unique values in selected variables
table(tab_act$vh13vision)

```

```{r}
# Ensure participant ID is kept
id_column <- "spid"

# Functional Limitations (Vision & Hearing)
functional_columns <- c("vh13vision", "vh13hearing", "vb135lglasses", "vb135lcontacts", "vb135lothvisaid")

# Readmission & Chronic Condition Care (searching for relevant terms)
readmission_columns <- grep("readmit|hospital", names(df), value = TRUE, ignore.case = TRUE)

# Caregiver Support & Digital Tools (searching for relevant terms)
caregiver_columns <- grep("care|support", names(df), value = TRUE, ignore.case = TRUE)

# Cognitive & Functional Ability Measures
cognitive_columns <- c("cb13onbspeed", "cb13onbstdev", "cb13onbacc", "cb13onbcorr", "cb13onberr", 
                       "cb13onbcomp", "cb13onbinteg", "cb13dcogbatm")

# Combine all selected columns
selected_columns <- unique(c(id_column, functional_columns, readmission_columns, caregiver_columns, cognitive_columns))


# Filter the dataset to keep only relevant columns
tab_act_filtered <- tab_act %>% select(all_of(selected_columns))

# View the filtered dataset
head(tab_act_filtered)
```

```{r}
# Rename columns to be more understandable
tab_act_filtered <- tab_act_filtered %>%
  rename(
    participant_id = spid,
    Vision_Status = vh13vision,
    Hearing_Status = vh13hearing,
    Uses_Glasses = vb135lglasses,
    Uses_Contacts = vb135lcontacts,
    Uses_Other_Visual_Aid = vb135lothvisaid,
    Cognitive_Speed = cb13onbspeed,
    Cognitive_Std_Dev = cb13onbstdev,
    Cognitive_Accuracy = cb13onbacc,
    Cognitive_Correct_Responses = cb13onbcorr,
    Cognitive_Errors = cb13onberr,
    Cognitive_Completion = cb13onbcomp,
    Cognitive_Integration = cb13onbinteg,
    Cognitive_Battery_Measure = cb13dcogbatm
  )

# View the renamed dataset
head(tab_act_filtered)
```


```{r}
# Define the relevant columns
selected_columns <- c("spid", "r13panel", "r13status", "r13casestdtmt", "r13casestdtyr",
                      "r13spstat1", "r13spstat2", "r13fqstat",
                      "r12status", "r11status")

# Filter the dataset to keep only relevant columns
tracker_data_filtered <- tracker_data %>% select(all_of(selected_columns))

# Rename the columns to meaningful names
tracker_data_filtered <- tracker_data_filtered %>%
  rename(
    participant_id = spid,
    Panel_ID = r13panel,
    Overall_Case_Status = r13status,
    Case_Status_Month = r13casestdtmt,
    Case_Status_Year = r13casestdtyr,
    Participant_Interview_Status_1 = r13spstat1,
    Participant_Interview_Status_2 = r13spstat2,
    Facility_Questionnaire_Status = r13fqstat,
    Previous_Year_Status = r12status,
    Two_Years_Ago_Status = r11status
  )

# View renamed dataset
head(tracker_data_filtered)

# We are using tracker data because it helps remove ineligible cases (e.g., deceased participants or refusals)
```

```{r}
# Merge datasets using left joins on 'participant_id'
merged_digital_tools <- tab_act_filtered %>%
  left_join(tracker_data_filtered, by = "participant_id")

# View summary of merged dataset
glimpse(merged_digital_tools)

# Save the final merged dataset as a CSV file
write_csv(merged_digital_tools, "~/desktop/Digital-tools_caregivers/NHATS_Merged_Digital_Tools.csv")
```

# Next, working on the remaining datasets that focus more on caregivers, patients, and merge them with the cleaned digital tools dataset:
```{r}
# Load libraries
library(haven)      # To read .sas7bdat files
library(tidyverse)  # For data wrangling
library(janitor)    # For cleaning column names
library(skimr)      # For summarizing datasets

# Load NHATS data files
nhats_sp <- read_sas("~/desktop/Digital-tools_caregivers/NHATS_R13_Final_Release_SAS/SP_files/NHATS_Round_13_SP_File.sas7bdat") # Sample Person Data
nhats_op <- read_sas("~/desktop/Digital-tools_caregivers/NHATS_R13_Final_Release_SAS/OP_files/NHATS_Round_13_OP_File.sas7bdat") # Caregiver Data
merged_digital_tools <- read_csv("~/desktop/Digital-tools_caregivers/NHATS_Merged_Digital_tools.csv") # Digital tool use data
inc_path <- read_sas("~/desktop/Digital-tools_caregivers/NHATS_R13_Final_Release_SAS/Inc_files/NHATS_R13_Int_Inc_Imp_File.sas7bdat")

# Clean column names
nhats_sp <- nhats_sp %>% clean_names()
nhats_op <- nhats_op %>% clean_names()
merged_digital_tools <- merged_digital_tools %>% clean_names()
nhats_inc <- inc_path %>% clean_names()  # Income & Socioeconomic Data

# Check variables
# colnames(nhats_sp)
# colnames(nhats_op)
# colnames(merged_digital_tools)
# colnames(nhats_inc)
```

```{r}
# Save the final merged dataset as a CSV file
# write_csv(nhats_op, "nhats_op.csv")
# write_csv(nhats_sp, "nhats_sp.csv")
# write_csv(inc_path, "inc_path.csv")
```

# Income dataset

```{r}
# Define the relevant columns
selected_columns_inc_path <- c("spid", "ia13toincimif", "ia13dtoincimi1", 
                               "ia13dtoincimi2", "ia13dtoincimi3", 
                               "ia13dtoincimi4", "ia13dtoincimi5", "ia13dtoincimreas")

# Filter the dataset to keep only relevant columns
inc_data_filtered <- nhats_inc %>% select(all_of(selected_columns_inc_path))

# View filtered dataset
head(inc_data_filtered)

# Rename the columns to meaningful names
inc_data_filtered <- inc_data_filtered %>%
  rename(
    participant_id = spid,
    Total_Imputed_Income = ia13toincimif,
    Income_Source_1 = ia13dtoincimi1,
    Income_Source_2 = ia13dtoincimi2,
    Income_Source_3 = ia13dtoincimi3,
    Income_Source_4 = ia13dtoincimi4,
    Income_Source_5 = ia13dtoincimi5,
    Missing_Income_Reason = ia13dtoincimreas
  )

# View renamed dataset
head(inc_data_filtered)
```

#SP dataset

```{r}
# Select and rename relevant variables from NHATS_SP (Older Adults Data)
nhats_sp_filtered <- nhats_sp %>%
  select(
    spid,              # Unique respondent ID
    r13dresid,         # Residential status (community, assisted living, nursing home)
    r13dgender,        # Gender of respondent
    r13d2intvrage,     # Age of respondent
    hc13disescn1, hc13disescn2, hc13disescn3, hc13disescn4, hc13disescn5, 
    hc13disescn6, hc13disescn7, hc13disescn8, hc13disescn9, hc13disescn10,  # Chronic diseases
    is13proxlivsp,     # Does caregiver live with respondent?
    is13prxyrelat,     # Relationship of the proxy caregiver
    is13prxygendr,     # Gender of the proxy caregiver
    em13paydevce1:em13paydevce6,  # Assistive technology usage
    te13intrntmd2, te13intrntmd3, te13intrntmd4, 
    te13computer, te13tablet, te13emailtext, te13online, 
    te13shoponli1, te13shoponli2, te13shoponli3, te13socialnet,
#Post acute care related
    hc13hosptstay, hc13hosovrnht,
# NEW: cognitive and functional decline variables
    cp13chgthink1:cp13chgthink8,  # Changes in thinking/memory
    mo13bedslf, mo13bedwout,      # Bed mobility
    rh13impactiv, rh13impcomp, rh13imphh,  # Limitations in daily life
    rh13funcback, rh13funcknees, rh13funcfeet, rh13funcwrist, # Body part limitations
    pc13walk6blks, pc13up10stair, pc13car20pnds # Mobility
  ) %>%
  
  rename(
    participant_id = spid,
    residence_type = r13dresid,         # Residential status (community, assisted living, nursing home)
    respondent_gender = r13dgender,     # Gender of respondent
    respondent_age = r13d2intvrage,     # Age of respondent
    has_heart_disease = hc13disescn1,   
    has_hypertension = hc13disescn2,
    has_lung_disease = hc13disescn3,
    has_diabetes = hc13disescn4,
    has_cancer = hc13disescn5,
    has_stroke = hc13disescn6,
    has_arthritis = hc13disescn7,
    has_dementia = hc13disescn8,
    has_depression = hc13disescn9,
    has_other_chronic = hc13disescn10,
    caregiver_lives_with_respondent = is13proxlivsp, # Does the caregiver live with respondent?
    caregiver_relationship = is13prxyrelat, # Relationship of the proxy caregiver
    caregiver_gender = is13prxygendr, # Gender of the proxy caregiver
    vision_aids = em13paydevce1, # Glasses, magnifiers
    hearing_aids = em13paydevce2, # Hearing aids
    cane = em13paydevce3, # Cane
    walker = em13paydevce4, # Walker
    wheelchair = em13paydevce5, # Wheelchair
    other_mobility_aid = em13paydevce6, # Grab bars, raised toilet seats
    telehealth_use = te13intrntmd2,         # Used internet for telehealth
    insurance_info_online = te13intrntmd3,  # Looked up insurance info online
    health_info_online = te13intrntmd4,     # Searched for health information online
    has_computer = te13computer,            # Owns and uses a computer
    has_tablet = te13tablet,                # Owns and uses a tablet
    uses_email_or_text = te13emailtext,     # Uses email or text messaging
    uses_online_services = te13online,      # Uses a computer for online tasks
    orders_groceries_online = te13shoponli1,# Orders groceries online
    online_banking = te13shoponli2,         # Uses online banking
    orders_prescriptions_online = te13shoponli3, # Orders prescription refills online
    social_media_use = te13socialnet,        # Uses social networking sites
    had_hosp_stay_12mo = hc13hosptstay, 
    num_hosp_stays = hc13hosovrnht, 
    
    # Cognitive change indicators
    change_memory = cp13chgthink1,
    change_decisionmaking = cp13chgthink2,
    change_following_instructions = cp13chgthink3,
    change_concentration = cp13chgthink4,
    change_language = cp13chgthink5,
    change_organization = cp13chgthink6,
    change_multitasking = cp13chgthink7,
    change_recognition = cp13chgthink8,

    # Functional mobility
    can_get_out_of_bed_alone = mo13bedslf,
    can_get_out_of_bed_with_aid = mo13bedwout,

    # Activity limitations
    needs_help_activities = rh13impactiv,
    needs_help_comprehension = rh13impcomp,
    needs_help_household = rh13imphh,

    # Physical function impairments
    back_limitation = rh13funcback,
    knee_limitation = rh13funcknees,
    foot_limitation = rh13funcfeet,
    wrist_limitation = rh13funcwrist,

    # Strength/mobility tests
    walk_six_blocks = pc13walk6blks,
    climb_stairs = pc13up10stair,
    lift_20_pounds = pc13car20pnds
  ) %>%
   mutate(
    residence_type = case_when(
      residence_type == 1 ~ "Community",
      residence_type == 2 ~ "Residential Care",
      residence_type == 3 ~ "Nursing Home",
      residence_type == 4 ~ "Assisted Living",
      TRUE ~ NA_character_
    ),
    respondent_gender = case_when(
      respondent_gender == 1 ~ "Male",
      respondent_gender == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    respondent_age = case_when(
      respondent_age == 1 ~ "65-69",
      respondent_age == 2 ~ "70-74",
      respondent_age == 3 ~ "75-79",
      respondent_age == 4 ~ "80-84",
      respondent_age == 5 ~ "85-89",
      respondent_age == 6 ~ "90+",
      TRUE ~ NA_character_
    ),
    caregiver_lives_with_respondent = case_when(
      caregiver_lives_with_respondent == 1 ~ "Yes",
      caregiver_lives_with_respondent == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    caregiver_relationship = case_when(
      caregiver_relationship == 1 ~ "Spouse/Partner",
      caregiver_relationship == 2 ~ "Child",
      caregiver_relationship == 3 ~ "Sibling",
      caregiver_relationship == 4 ~ "Other Relative",
      caregiver_relationship == 5 ~ "Friend/Neighbor",
      caregiver_relationship == 6 ~ "Paid Caregiver",
      caregiver_relationship == 7 ~ "Other Non-Relative",
      caregiver_relationship == 8 ~ "Unknown",
      TRUE ~ NA_character_
    ),
    caregiver_gender = case_when(
      caregiver_gender == 1 ~ "Male",
      caregiver_gender == 2 ~ "Female",
      TRUE ~ NA_character_
    ),
    # Convert assistive technology usage (1=Yes, 2=No, -9=Missing)
    across(starts_with("vision_aids"):starts_with("other_mobility_aid"), ~ case_when(
      . == 1 ~ "Yes",
      . == 2 ~ "No",
      TRUE ~ "Missing"
    )),
    # Convert digital tool usage (1=Yes, 2=No, -9=Missing)
    across(starts_with("telehealth_use"):starts_with("social_media_use"), ~ case_when(
      . == 1 ~ "Yes",
      . == 2 ~ "No",
      TRUE ~ "Missing"
    )),
     # Existing binary conversions
  had_hosp_stay_12mo_bin = case_when(
    had_hosp_stay_12mo == 1 ~ 1,
    had_hosp_stay_12mo == 2 ~ 0,
    TRUE ~ NA_real_
  ),
  multiple_hosp_stays = case_when(
    num_hosp_stays >= 2 ~ 1,
    num_hosp_stays < 2 ~ 0,
    TRUE ~ NA_real_
  ),

  # New: Categorized version of number of hospital stays
  hosp_stay_category = case_when(
    num_hosp_stays == 0 ~ "None",
    num_hosp_stays == 1 ~ "One",
    num_hosp_stays >= 2 ~ "Multiple",
    TRUE ~ NA_character_
  )
  )

write_csv(nhats_sp_filtered, "~/desktop/Digital-tools_caregivers/nhats_sp_filtered.csv")
```

#OP dataset

```{r}
# Select and rename relevant variables from NHATS_OP (Caregivers Data)
nhats_op_filtered <- nhats_op %>%
  select(
    spid,             # Unique respondent ID (for merging)
    op13relatnshp,    # Relationship of caregiver to respondent
    op13proxy,        # Whether caregiver answered on behalf of respondent
    op13dage,         # Age of the caregiver
    op13numhrsday,    # Hours of caregiving per day
    op13numdayswk     # Days of caregiving per week
  ) %>%
  rename(
    participant_id = spid,
    caregiver_relationship_to_respondent = op13relatnshp, # Relationship of caregiver to respondent
    caregiver_answered_for_respondent = op13proxy, # Whether caregiver answered on behalf of respondent
    caregiver_age = op13dage, # Age of the caregiver
    caregiver_hours_per_day = op13numhrsday, # Hours of caregiving per day
    caregiver_days_per_week = op13numdayswk # Days of caregiving per week
  )%>%
  mutate(
    caregiver_relationship_to_respondent = case_when(
      caregiver_relationship_to_respondent == 1 ~ "Spouse/Partner",
      caregiver_relationship_to_respondent == 2 ~ "Child",
      caregiver_relationship_to_respondent == 3 ~ "Sibling",
      caregiver_relationship_to_respondent == 4 ~ "Other Relative",
      caregiver_relationship_to_respondent == 5 ~ "Friend/Neighbor",
      caregiver_relationship_to_respondent == 6 ~ "Paid Caregiver",
      caregiver_relationship_to_respondent == 7 ~ "Other Non-Relative",
      caregiver_relationship_to_respondent == 8 ~ "Unknown",
      TRUE ~ NA_character_
    ),
    
    caregiver_answered_for_respondent = case_when(
      caregiver_answered_for_respondent == 1 ~ "Yes",
      caregiver_answered_for_respondent == 2 ~ "No",
      TRUE ~ NA_character_
    ),
    
    caregiver_age = case_when(
      caregiver_age == -9 ~ NA_real_,  # Replace missing values
      TRUE ~ caregiver_age  # Keep actual age values
    ),
    
    caregiver_hours_per_day = case_when(
      caregiver_hours_per_day == -9 ~ NA_real_,
      caregiver_hours_per_day == -1 ~ NA_real_,
      TRUE ~ caregiver_hours_per_day
    ),
    
    caregiver_days_per_week = case_when(
      caregiver_days_per_week == -9 ~ NA_real_,
      caregiver_days_per_week == -1 ~ NA_real_,
      TRUE ~ caregiver_days_per_week
    )
  )
```



Merge Caregiver & Digital Tool Use Data

Merge NHATS caregiving data with digital tool usage

```{r}
# Merge NHATS_SP with NHATS_OP (Caregiver + Care Recipient Data)
nhats_merged <- nhats_sp_filtered %>%
  left_join(nhats_op_filtered, by = "participant_id") %>%
  left_join(merged_digital_tools, by = "participant_id") %>%
  left_join(inc_data_filtered, by = "participant_id")

# Check merged dataset
glimpse(nhats_merged)

write_csv(nhats_merged, "~/desktop/Digital-tools_caregivers/nhats_merged.csv")
```






# Work on the NHATS merged dataset to deal with missingness and make it ready to merge with HRRP dataset:

# Handling Missing Values

```{r}
# Check for missing values in each column
colSums(is.na(nhats_merged))  # Shows number of missing values per column
```

#Handling Missing Categorical Variables
#Categorical values -> "Unknown" for missing values
```{r}
# List of categorical columns to replace NAs with "Unknown"
categorical_cols <- c("residence_type", "caregiver_lives_with_respondent", 
                      "caregiver_relationship", "caregiver_gender",
                      "caregiver_relationship_to_respondent", "caregiver_answered_for_respondent")

# Replace missing values with "Unknown"
nhats_merged[categorical_cols] <- lapply(nhats_merged[categorical_cols], function(x) ifelse(is.na(x), "Unknown", x))
```

#Handling Missing Numeric Variables
```{r}
# Handling Missing Values in Age Variables 
# Keep respondent_age as a character (preserving the range)
nhats_merged$respondent_age <- as.character(nhats_merged$respondent_age)

# keep missing ages them as NA
nhats_merged$respondent_age[nhats_merged$respondent_age == ""] <- NA  # Ensures empty strings are NA
```


```{r}
# Handling Caregiver Age (Numeric) 
# Convert caregiver_age to numeric and replace NA with the median age
nhats_merged$caregiver_age <- as.numeric(nhats_merged$caregiver_age)
nhats_merged$caregiver_age[is.na(nhats_merged$caregiver_age)] <- median(nhats_merged$caregiver_age, na.rm = TRUE)
```


```{r}
# Handling Missing Values for Care Hours and Days
nhats_merged$caregiver_hours_per_day[is.na(nhats_merged$caregiver_hours_per_day)] <- median(nhats_merged$caregiver_hours_per_day, na.rm = TRUE)
nhats_merged$caregiver_days_per_week[is.na(nhats_merged$caregiver_days_per_week)] <- median(nhats_merged$caregiver_days_per_week, na.rm = TRUE)

print(colSums(is.na(nhats_merged)))
```


# Handling Special Cases (Total_Imputed_Income has -1)
# Replacing -1 with NA (Your code)
```{r}
# Replace -1 with NA in Total_Imputed_Income
nhats_merged$Total_Imputed_Income[nhats_merged$Total_Imputed_Income == -1] <- NA

head(nhats_merged)
```

```{r}
#Standardizing Binary Variables 
binary_cols <- c("has_heart_disease", "has_hypertension", "has_lung_disease", "has_diabetes",
                 "has_cancer", "has_stroke", "has_arthritis", "has_dementia", "has_depression", "has_other_chronic")

nhats_merged[binary_cols] <- lapply(nhats_merged[binary_cols], function(x) {
  x <- as.numeric(x)  # Ensure numeric
  ifelse(x == 1, 1, ifelse(x == 2, 0, NA))  # 1 = Yes, 0 = No, NA for other values
})
```

```{r}
# Convert income source columns to numeric
income_cols <- c("Income_Source_1", "Income_Source_2", "Income_Source_3", "Income_Source_4", "Income_Source_5")
nhats_merged[income_cols] <- lapply(nhats_merged[income_cols], as.numeric)
```

```{r}
# Handling Duplicates ----
# Check for duplicate participant IDs
duplicate_count <- sum(duplicated(nhats_merged$participant_id))
print(paste("Number of duplicate participant IDs:", duplicate_count))

# If duplicates exist, remove all but the first occurrence
nhats_merged <- nhats_merged[!duplicated(nhats_merged$participant_id), ]
```

```{r}
# Encoding Categorical Variables 
# Convert categorical variables to factors
categorical_cols <- c("residence_type", "caregiver_relationship", "caregiver_gender")
nhats_merged[categorical_cols] <- lapply(nhats_merged[categorical_cols], as.factor)
```

```{r}
#Removing or Transforming Outliers 
# Identify outliers in caregiver_hours_per_day using IQR
Q1 <- quantile(nhats_merged$caregiver_hours_per_day, 0.25, na.rm = TRUE)
Q3 <- quantile(nhats_merged$caregiver_hours_per_day, 0.75, na.rm = TRUE)
IQR_value <- Q3 - Q1
upper_bound <- Q3 + 1.5 * IQR_value
lower_bound <- Q1 - 1.5 * IQR_value

# Replace outliers with the median
median_caregiver_hours <- median(nhats_merged$caregiver_hours_per_day, na.rm = TRUE)
nhats_merged$caregiver_hours_per_day[nhats_merged$caregiver_hours_per_day > upper_bound | nhats_merged$caregiver_hours_per_day < lower_bound] <- median_caregiver_hours
```

```{r}
# Verifying Logical Consistency
# If a participant has no caregiver, ensure caregiver_hours_per_day is 0 or NA
nhats_merged$caregiver_hours_per_day[nhats_merged$caregiver_relationship == "No caregiver"] <- 0

# Final check for missing values
colSums(is.na(nhats_merged))
```

```{r}
write_csv(nhats_merged, "~/desktop/Digital-tools_caregivers/Updated_nhats_merged.csv")
```




